---
title: "Basic Inferential Data Analysis"
subtitle: "Statistical Inference Course Project: Part 2"
author: "Harkishan Grewal"
date: "11/05/2020"
output: pdf_document
---

## Overview

Analyze the *ToothGrowth* data in the R datasets package.

1. Load the *ToothGrowth* data and perform some basic exploratory data analyses
2. Provide a basic summary of the data
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by 
supp and dose
4. State conclusions and assumptions needed for the conclusions


## Analysis

```{r, echo=FALSE}
library("ggplot2")
```

### *ToothGrowth* data
```{r}
# Load the ToothGrowth data
data("ToothGrowth")
str(ToothGrowth)
```


#### Description

The response is the length of odontoblasts (cells responsible for tooth growth) 
in 60 guinea pigs. Each animal received one of three dose levels of vitamin C 
(0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic 
acid (a form of vitamin C and coded as VC).

### Basic Summary
```{r}
summary(ToothGrowth)
```

```{r}
g <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) 
g <- g + geom_boxplot(aes(fill = factor(dose)))
g <- g + facet_grid(. ~ supp)
g <- g + ggtitle("Dose Amount vs Tooth Length")
g <- g + xlab("Dose Amount") + ylab("Length")
g
```

### Hypothesis Test

Now tooth growth by supplement using a t-test will be compared.

```{r}
t.test(len ~ supp, ToothGrowth)
```

The p-value of this test was 0.06063.

Since the p-value is greater than 0.05 and the confidence interval of the test 
contains zero, it can be concluded that supplement types seem to have no 
impact on tooth growth based on this test.

Now a comparison of tooth growth by dose will be done, looking at the 
different pairs of dose values.

```{r}
t.test(len ~ dose, ToothGrowth[ToothGrowth$dose %in% c(0.5, 1), ])
```

```{r}
t.test(len ~ dose, ToothGrowth[ToothGrowth$dose %in% c(1, 2), ])
```

The p-value of each test was essentially zero and the confidence interval of 
each test does not cross over zero.

Based on this result, it can be assumed that the average tooth length increases 
with an increasing dose, and therefore the null hypothesis can be rejected.

### Conclusions and Assumptions

Based on the t-test analysis, it can be concluded that supplement 
type has no effect on tooth growth, and increasing the dose level leads to 
increased tooth growth.