---
title: "Simulation Exercise"
subtitle: "Statistical Inference Course Project: Part 1"
author: "Harkishan Grewal"
date: "11/05/2020"
output: pdf_document
---

## Overview

In this project, I will investigate the exponential distribution in R and 
compare it with the Central Limit Theorem. The exponential distribution can be 
simulated in R with rexp(n, lambda) where lambda is the rate parameter. The 
mean of exponential distribution is 1/lambda and the standard deviation is also 
1/lambda. Set lambda = 0.2 for all of the simulations. I will investigate the 
distribution of averages of 40 exponentials.

## Simulations

Simulate the distribution of 1000 averages of 40 random exponentials.

```{r, echo=FALSE}
library("ggplot2")
```

```{r}
# set seed for reproducability
set.seed(1000)

# number of exponentials
n = 40

# Set lambda = 0.2 for all of the simulations
lambda = 0.2

# number of simulations
samples <- 1000

# do 1000 random simulations
mns = NULL

for (i in 1 : samples) {
    mns = c(mns, mean(rexp(n, lambda)))
}
```

## Sample Mean versus Theoretical Mean

```{r, warning=FALSE, message=FALSE}
# means of simulations
sample_mean <- mean(mns)

# theoretical mean
theoretical_mean <- 1 / lambda

df <- data.frame(mns)

g <- ggplot(df, aes(x = mns))
g <- g + geom_histogram(aes(y = ..density..), fill = "blue", color = "grey")
g <- g + xlim(range(density(mns)$x))
g2 <- g + geom_vline(
    aes(xintercept = sample_mean, color="sample mean"),
    lwd = 1)
g2 <- g2 + geom_vline(
    aes(xintercept = theoretical_mean, color="theoretical mean"))
g2 <- g2 + labs(
    title = "Sample Mean vs Theoretical Mean", x = "Mean", y = "Density")
g2
```

```{r}
# Print Sample Mean and Theoretical Mean
c(sample_mean, theoretical_mean)
```

The sample mean is `r sample_mean` which is close to the theoretical mean of 
`r theoretical_mean`.

## Sample Variance versus Theoretical Variance

```{r}
# variance of simulations
sample_var <- var(mns)

# theoretical variance
theoretical_var <- (1 / lambda)^2 / n

# Print Sample Variance and Theoretical Variance
c(sample_var, theoretical_var)
```

The sample variance is `r sample_var` which is close to the theoretical 
variance of `r theoretical_var`.

## Distribution

```{r, message=FALSE, warning=FALSE}
# theoretical standard deviation
theoretical_sd <- 1 / (lambda * sqrt(n))
g3 <- g + geom_density(aes(y = ..density..), color = "red", lwd = 1.0)
g3 <- g3 + stat_function(
    fun = dnorm, args = list(mean = theoretical_mean, sd = theoretical_sd),
    color = "blue", lwd = 1.0)
g3 <- g3 + labs(title = "Distribution", x = "Mean", y = "Density")
g3
```

The distribution (red line) is approximately normal (blue line).
