---
title: "Automatic vs Manual: Which Transmission Has Better MPG?"
subtitle: "Regression Models Course Project"
author: "Harkishan Grewal"
date: "12/05/2020"
output: pdf_document
---

## Executive Summary

This project uses the *mtcars* data set to examine the relationship between 
miles-per-gallon (MPG) and other variables. The project attempts to determine 
which transmission (automatic or manual) is better for MPG, and quantifies the 
difference.

The Regression Models section draws inference with a simple linear regression 
model and a multiple regression model. Both models show that manual 
transmission cars, on average, have a significantly higher MPG than automatic 
transmission cars.

In the linear model, the mean difference is 7.245 MPG; the average MPG for 
automatic transmission cars is 17.147 MPG, and the average MPG for manual 
transmission cars is 24.392 MPG. In the multiple regression model, the MPG 
difference is 2.9358 MPG.

The linear model shows that transmission type only accounts for 36% of the MPG
performance. The multivariable regression model the other variables explain
85% of the MPG performance.

## Exploratory Analysis and Visualizations

```{r}
# Load Datasets
library(datasets)
data(mtcars)

# Get overview of dataset
head(mtcars)

# Get overview of Variables
str(mtcars)

# Statistical Summary of MPG variable
summary(mtcars$mpg)

# Automatic vs Manual Transmission - Boxplot
library(ggplot2)
mtcars$am <- as.factor(mtcars$am)
trans_type <- ggplot(aes(x=am, y=mpg), data=mtcars) + geom_boxplot(aes(fill=am))
trans_type <- trans_type + labs(title = "Automatic vs Manual Transmission - Boxplot")
trans_type <- trans_type + xlab("Transmission Type")
trans_type <- trans_type + ylab("MPG")
trans_type <- trans_type + labs(fill = "Legend \n0 = Automatic\n1 = Manual")
trans_type

# Automatic vs Manual Transmission - Hypothesis Test:
auto <- mtcars[mtcars$am == "0",]
man <- mtcars[mtcars$am == "1",]
t.test(auto$mpg, man$mpg)
```


## Regression Models

### Linear Regression
```{r}
lrModel <- lm(mpg ~ am, data = mtcars)
summary(lrModel)
```

### Multivariable Regression
```{r}
mrModel <- lm(mpg~am + cyl + hp + wt, data = mtcars)
summary(mrModel)
```

### Linear vs Multivariable Regression
```{r}
anova(lrModel, mrModel)
```


## Conclusion
A linear regression model is used to determine the effect of transmission type 
on gas mileage (MPG). The model shows that manual transmission cars provide 
7.25MPG (the am1 coefficient) better performance than automatic cars. Based on 
the R-squared value, transmission type only accounts for 36% of the MPG 
performance. Thus, a linear regression model is not sufficient to 
answer Motor Trend’s questions with high certainty.

A multivariable regression model is used to account for the effect other 
variables on gas mileage (e.g. number of cylinders, engine horsepower, 
vehicle weight, etc.) This model showed a 1.48MPG increase from manual 
transmission cars over automatic cars with the additional variables explaining 
85% of the MPG performance.

## Appendix
```{r}
# Scatterplot matrix of the mtcars dataset
pairs(mpg ~ ., data = mtcars)

# Scatterplots of the multivariable regression model residuals
par(mfrow = c(2,2))
#plot(mrModel)
```